---
title: "Programmation Stochastique avec Recours"
author: "P. Hénaff"
date: "Version: `r format(Sys.Date(), '%d %b %Y')`"
output:
  beamer_presentation:
    colortheme: dolphin
    theme: Montpellier
  slidy_presentation: default

header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{graphicx}
  - \usepackage{subfig}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{xfrac}
  - \linespread{1.05}
  - \usepackage[utf8]{inputenc}
  - \usepackage{pgfplots}
  - \usepackage{tikz}
  - \usetikzlibrary{shapes}
  - \usetikzlibrary{external}
  - \usepgfplotslibrary{external}
  - \usepackage[ruled, vlined, linesnumbered]{algorithm2e}


bibliography: ../library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```




```{r, results='hide', message=FALSE}
  library(foreach)
  library(doParallel)
  library(kableExtra)
  library(linprog)
  library(gtools)
  library(pracma)
  library(kernlab)
  library(optiSolve)
  library(piqp)
  library(latex2exp)
```

## Arbre de décision

\scalebox{.7}{
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=8cm},
  level 2/.style={sibling distance=4cm},
  level 3/.style={sibling distance=2cm}]
  \node {1}
    child {node {2 high}
      child {node {4 high}
        child {node {S-1}}
        child {node {S-2}}
        }
      child {node {5 low}
        child {node {S-3}}
        child {node {S-4}}
      }
    }
    child {node {3 low}
      child {node {6 high}
        child {node {S-5}}
        child {node {S-6}}
        }
      child {node {7 low}
        child {node {S-7}}
        child {node {S-8}}
      }
    };
\end{tikzpicture}
}


## Solution


```{r, echo=FALSE}
Amat <- matrix(0, nrow=15, ncol=30)

sce_up = c(1.25, 1.14)
sce_down = c(1.06, 1.12)
sce = rbind(sce_up, sce_down)

Amat[1,1:2] = 1
Amat[2,1:2] = -sce_up; Amat[2,3:4] = 1
Amat[3,1:2] = -sce_down; Amat[3,5:6] = 1
Amat[4,3:4] = -sce_up; Amat[4,7:8] = 1
Amat[5,3:4] = -sce_down; Amat[5,9:10] = 1
Amat[6,5:6] = -sce_up; Amat[6,11:12] = 1
Amat[7,5:6] = -sce_down; Amat[7,13:14] = 1

Amat[8,7:8] = sce_up; Amat[8,15:16] = c(-1, 1)
Amat[9,7:8] = sce_down; Amat[9,17:18] = c(-1, 1)
Amat[10,9:10] = sce_up; Amat[10,19:20] = c(-1, 1)
Amat[11,9:10] = sce_down; Amat[11,21:22] = c(-1, 1)
Amat[12,11:12] = sce_up; Amat[12,23:24] = c(-1, 1)
Amat[13,11:12] = sce_down; Amat[13,25:26] = c(-1, 1)
Amat[14,13:14] = sce_up; Amat[14,27:28] = c(-1, 1)
Amat[15,13:14] = sce_down; Amat[15,29:30] = c(-1, 1)

bvec = matrix(0, nrow=15, ncol=1)

W.0 = 55000
G = 80000
r_short = 4/100
r_long = 1/100
prob = 1/8

bvec[1,1] = W.0
bvec[8:15,1] = G
cvec = matrix(0, nrow=1, ncol=30)
cvec[1, seq(15,30,by=2)] = r_long * prob
cvec[1, seq(16,30,by=2)] = -r_short * prob

bvec = as.vector(bvec)
cvec = as.vector(cvec)

res = solveLP(cvec, bvec, Amat, const.dir=rep("==", length(bvec)), lpSolve=TRUE, maximum=TRUE,verbose=1)

sol = matrix(0, nrow=7, ncol=3)
for(i in seq(7)) {
  sol[i,] = c(i, res$solution[(2*i-1):(2*i)])
}
colnames(sol) <- c("Node", "Stock", "Bond")
```

```{r, echo=FALSE}
kable(sol, caption="Optimal asset allocation at each node in the 3-period tree",
booktabs=T, digits=0) %>%
kable_styling(latex_options="hold_position")
```

## Surplus or Shortfall by scenario

```{r es, echo=FALSE}
wealth <- matrix(0, nrow=8, ncol=3)
colnames(wealth) <- c("Scenario", "Shortfall", "Excess")
for(i in seq(8)) {
  wealth[i,] = c(i, res$solution[14+2*i], res$solution[14+2*i-1])
}
kable(wealth, caption="Shortfall or excess by scenario",
booktabs = T, digits=0)
```


## Stochastic vs Myopic Solution

```{r, echo=FALSE}
return.index <- permutations(2,3,c(0,1), repeats.allowed = T)
return.high <- 1.25
return.low <- 1.06

idx <- rowSums(return.index)
return.by.scenario <- return.high ^ (3-idx) * return.low ^ idx
wealth.by.scenario <- W.0 * return.by.scenario
```

```{r ev, echo=FALSE}
options(knitr.kable.NA = "")
df = data.frame(Scenarios=seq(8),
W1.Short=-pmin(wealth.by.scenario-G, 0),
W1.Long = pmax(wealth.by.scenario-G, 0),
W2.Short=wealth[,2],
W2.Long=wealth[,3])
kable(replace(df, df==0, NA), col.names=c("Scenario", "Shortfall", "Surplus", "Shortfall", "Surplus"),
booktabs=T, digits=0,
caption="Expected Value vs. Recourse solutions of the asset allocation problem.") %>%
  kable_paper() %>%
  add_header_above(c(" "=1, "EV solution"=2, "Recourse solution"=2)
  )
```



```{r, echo=F}
expected.cost.ev = mean( -df$W1.Short*r_short + df$W1.Long*r_long)
expected.cost.sp = mean( -df$W2.Short*r_short + df$W2.Long*r_long)
VSS = expected.cost.sp - expected.cost.ev
```

## Augmented Lagrangian

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{$\rho$, tol, $k_{\mbox{max}}$}
\Output{$x^*$}
\BlankLine
$k \leftarrow 0;\ \  \lambda_\leftarrow 0$\;
\While{$k<k_{\mbox{max}}$ and $t > tol$}{
    $x_{k+1} \leftarrow  \underset{x}{\mathrm{argmin}} \phi_k(x)$\;
    $\lambda_{k+1} \leftarrow \lambda_k - \rho g(x_{k+1})$\;
    $k \leftarrow k+1$\;
    $t_1  \leftarrow ||\lambda_{k+1} - \lambda_k||$\;
    $t_2 \leftarrow ||g(x_{k+1}||$\;
    $t \leftarrow \max(t_1, t_2)$\;
    }
\caption{Augmented Lagrangian algorithm}
\end{algorithm}


## Algo Progressive Hedging

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{$\rho$, tol, $i_{\mbox{max}}$}
\Output{$x^*$}
\BlankLine
$i \leftarrow 0; \lambda_i \leftarrow 0; \mbox{converged} \leftarrow False$\;
\While{$k<k_{\mbox{max}}$ and $! \mbox{converged}$}{
    Solve the $K$ subproblems  to obtain $x^{i+1}_k, k = 1, \ldots, K$\;
    Compute  $\hat{x}^{i+1} = \sum_{k=1}^K \pi_k x^{i+1}_k$\;
    Update the multipliers: $\lambda^{i+1}_k = \lambda^i_k - \rho(x^{i+1}_k - \hat{x}^{i+1})$\;
    $i \leftarrow i+1$\;
    $t_1  \leftarrow ||\lambda^{i+1} - \lambda^i|| < \mbox{tol}_1$\;
    $t_2 \leftarrow ||g^1(x^{i+1}|| < \mbox{tol} < \mbox{tol}_2$\;
    $t_3 \leftarrow ||g^2(x^{i+1}|| < \mbox{tol} < \mbox{tol}_3$\;
    $\mbox{converged} \leftarrow \max(t_1, t_2, t_3)$\;
    }
\caption{Progressive Hedging algorithm}
\end{algorithm}


```{r, echo=FALSE}
P <- matrix(0, nrow=4, ncol=4)
P[3,3] = 1
P[4,4] = 1
c <- rep(0, 4)
G <- matrix(c(1, 1, 0, 0,
              -1, -3, -1, 0,
              -4, -2, 0, -1), nrow=3, byrow=TRUE)
h <- c(10, -25, -25)
x_lb <- rep(0, 4)
sol <- solve_piqp(P=P, c=c, G=G, h=h, x_lb=x_lb )
```

```{r, echo=F}
sub.prob <- function(x.hat, r, w, q) {
P <- zeros(3,3)
P[1,1] <- r
P[2,2] <- r
P[3,3] <- 2
c <- c(w - r*x.hat, 0)
G <- matrix(c(1, 1, 0,
              -q[1], -q[2], -1), nrow=2, byrow = TRUE)
h <- c(10, -25)
x_lb <- rep(0,3)
sol <- solve_piqp(P=P, c=c, G=G, h=h, x_lb = x_lb)
sol
}
```


```{r, echo=F}
q = matrix(c(1,3,4,2), nrow=2, byrow=TRUE)
w = matrix(0, nrow=2, ncol=2)
x.hat <- c(0,0)
x = matrix(0, nrow=2, ncol=2)
tol <- 1.0e-2
x[1,] <- sub.prob(x.hat, r=0, w[1,], q[1,])$x[1:2]
x[2,] <- sub.prob(x.hat, r=0, w[2,], q[2,])$x[1:2]
x.hat <- colMeans(x)

conver <- FALSE
iter.count <- 0
iter.max <- 100
r = 2
x.iter <- matrix(0, nrow=iter.max, ncol=2)

while(!conver & (iter.count < iter.max)) {
iter.count <- iter.count + 1
x.iter[iter.count,] = x.hat
x[1,] <- sub.prob(x.hat, r, w[1,], q[1,])$x[1:2]
x[2,] <- sub.prob(x.hat, r, w[2,], q[2,])$x[1:2]


x.hat.old <- x.hat
x.hat <- colMeans(x)
w.old <- w
w[1,] = w[1,] + r * (x[1,] - x.hat)
w[2,] = w[2,] + r * (x[2,] - x.hat)

# convergence test
test.1 <- norm(x.hat-x.hat.old, type="2")
test.2 <- norm(w - w.old, type="2")

conver <- (test.1 <= tol) & (test.2 <= tol)
}
```

## First Example

```{r, PHA-plot, echo=FALSE, out.width="70%",fig.cap="Progressive Hedging Iterations", fig.align='left'}
plot(seq(iter.count),x.iter[1:iter.count,1], ylim=c(2, 8), type="l", col="blue", xlab="Iterations",
     ylab=expression(X))
lines(seq(iter.count), x.iter[1:iter.count,2], type="l", col="red")
abline(h=2.5)
abline(h=7.5)
```


## Second example: A 3-stage optimization problem

```{r, fig.align="center", out.width="100%", echo=FALSE}
knitr::include_graphics("Mutistage-Optimization-figure1.pdf")
```

## Linear Convergence of the Progressive Hedging Algorithm

```{r, fig.align="center", out.width="100%", echo=FALSE}
knitr::include_graphics("PHA-plot-1-1.pdf")
```
